Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-10-09 nicolas/PPIPL3/ReportV4.tex
Methodology update of validation

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. This paper evaluates predictions based on
three measures: Common Neighbors (\textbf{CN}), based on the count
of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
and the degree-normalized count of paths of length 3 (\textbf{L3}).
Furthermore

@@

@@

For achieving the described results, human and rice PPI networks are
used and compared using state-of-the-art methods, as well as the proposed
ones (CN, A3, L3). In the case of the human network, human interactome
(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
were used. A massive experimental assay was carried on and its results
were consolidated and used to build a validation network (\emph{HI-III}).

@@

For each network of protein interactions, the following procedure
was performed 10 times in order to address the stochastic nature of
the process and have a consensus:
\begin{itemize}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied.
\end{itemize}
It is worth mentioning that some exploration experiments were carried
out with interactions removal percentages of 2, 5, 10 and 20\%, as
well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
parameter set was selected because it either represented a marginal
gain (results not shown) or because it was a common parameter selection
in related literature.

@@

For the rice interactome, the different model-features combinations
were trained and validated. After executing the mentioned routines,
the results are shown in Figure \ref{F1}. First, one should have
a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
without any additional feature included. The plot below shows those
results, and one can see that its mean performance using the AUC metric
is $0.52$, and that the results among the 10 repetitions are consistent.
This results mean that the model using only the default features perform
barely as good as a random choice of the labels. This result can also
be assessed when looking at the confusion matrix, where a precision
of $0.5168$ and a recall of $0.4589$ can be derived.

@@

Next experiment carried out was to add each prediction method score
as a prediction feature, that is, use the score as the eleventh input
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
It can be observed again that all 10 experiments have small variability
among them and perform significantly better than the baseline, with
an area under the ROC curve of $0.90$, with small standard deviation.
When looking at the confusion matrix for this model, turns out that
most of the guesses are true positives and true negatives, resulting
in a precision of $0.9285$ and a recall of $0.7420$. Same analyses
are done with the count of paths of length 3 (\textbf{A3}) and with
the degree-normalized length-3 score (\textbf{L3}) and results are
presented in figures \ref{F3} and \ref{F4}, respectively, resulting
also in area under the curve of 0.9 for both cases.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2}Results for \texttt{Node2Vec} model with CN feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3}Results fir \texttt{Node2Vec} model with A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4}Results for \texttt{Node2Vec} model with L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
\end{figure}

The precision when using the count of paths of length 3 (A3) is $0.8587$
while the recall is $0.7685$. It is interesting to observe that precision
decreased $7.5\%$ when compared to the previous case, but recall
increased $3.6\%$. A similar situation is obtained for the model
with the normalized score (L3), whose precision decreased to $0.8936$
($-3.8\%$) and recall increased to $0.7610$ ($+2.6\%$). It can
be seen also that the ROC curve rapidly achieves higher values of
true positive rate in the A2-featured model when compared to the other
models.

Finally, an relevant evaluation on the models that include a handcrafted
feature is necessary: How important is the appended feature for the
model result? The answer can be observed by the feature importance
plots for CN, A3 and L3 in Figure \ref{F5-importance}. Each plot
resembles the number of times that each feature creates a bifurcation
in the underlying decision trees that \texttt{XGBoost} uses. The more
bifurcations, the higher the importance a feature has on the model
itself. All models have the appended feature as the most important
by a large margin.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F5-importance}Importance for \texttt{Node2Vec} models: with
L3 feature}
\par\end{centering}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot 1}
\par\end{raggedleft}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot 1}
\par\end{raggedleft}
\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot 1}
\end{figure}

@@

(PENDING FROM HERE)

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-II-14}}

@@

\begin{figure}[h]
\caption{Methods Comparison for \emph{HI-TESTED}}

@@

@@

Taking into account the different results validated in this report,
one can conclude that length-3 path methodologies might work better
on protein-protein interactions than its traditional length-2 (TCP
based) counterparts. On the other hand, it can be seen that degree-normalization
has little effect on the predictions, i.e., non-normalized $A{{}^3}$
matrix predictions are still a good methodology for edge prediction
on PPI networks.

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical paths of length 2 are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

\begin{comment}
JF: please update the conclusions - WSD does not seem relevant

NL: Done
\end{comment}

>>>>

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. This paper evaluates predictions based on
three measures: Common Neighbors (\textbf{CN}), based on the count
of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
and the degree-normalized count of paths of length 3 (\textbf{L3}).
Furthermore, the Machine Learning (ML) process is assessed for each
metric as the only feature, for the features obtained by the Node2Vec
algorithm and for a combination of the Node2Vec and

@@

NL: Done!

@@

For achieving the described results, human and rice PPI networks are
used and compared using state-of-the-art methods, as well as the proposed
ones (CN, A3, L3). In the case of the human network, human interactome
(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
were used. A massive experimental assay was carried on by other authors
and its results were consolidated and used to build a validation network
(\emph{HI-III}).

@@

For each network of protein interactions, the following procedure
was performed 10 times in order to address the stochastic nature of
the process and have a consensus:
\begin{itemize}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied in training,
but validation is performed using the remaining unbalanced data.
\end{itemize}
It is worth mentioning that some exploration experiments were carried
out with interactions removal percentages of 2, 5, 10 and 20\%, as
well as train-test partitions of 15-85, 80-20 and 75-25, and the explained
parameter set was selected because it either represented a marginal
gain (results not shown) or because it was a common parameter selection
in related literature.

@@

For the rice interactome, the different model-features combinations
were trained and validated. After executing the mentioned routines,
the results are shown in Figure \ref{F1}. First, one should have
a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
without any additional feature included. The plot below shows those
results, and one can see that its mean performance using the AUC metric
is $0.93$, and that the results among the 10 repetitions are consistent.
This results mean that the model using only the default features performs
very good. This result can also be assessed when looking at the confusion
matrix, where a precision of $0.9814$ and a recall of $0.8649$ can
be derived.

@@

Next experiment carried out was to add each prediction method score
as a prediction feature, that is, use the score as the eleventh input
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
It can be observed again that all 10 experiments have small variability
among them and perform better than the baseline, with an area under
the ROC curve of $0.98$, with small standard deviation. When looking
at the confusion matrix for this model, turns out that most of the
guesses are true positives and true negatives, resulting in a precision
of $0.9952$ and a recall of $0.9108$. Same analyses are done with
the count of paths of length 3 (\textbf{A3}) and with the degree-normalized
length-3 score (\textbf{L3}) and results are presented in figures
\ref{F3} and \ref{F4}, respectively, resulting also in area under
the curve of $0.97$ and $0.98$, respectively.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2}Results for \texttt{Node2Vec} with CN feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3}Results for \texttt{Node2Vec} with A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CMA3_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4}Results for \texttt{Node2Vec} with L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{ML_Metric/ROCL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{ML_Metric/CML3_SUMMARY}
\end{figure}

The precision when using the count of paths of length 3 (A3) is $0.9935$
while the recall is $0.9055$. It is interesting to observe that precision
decreased $0.16\%$ and recall decreased $0.58\%$. A similar situation
is obtained for the model with the normalized score (L3), whose precision
decreased to $0.9944$ ($-0.08\%$) and recall decreased to $0.8891$
($-2.38\%$). It can be seen also that the ROC curve achieves high
values of true positive rate sooner in the A2-featured model than
on the other models.

Another relevant assessment was carried out to verify the predictive
power of the metrics. \texttt{XGBoost} models were trained only using
each metric, for the same 10 experiments. Results are shown in Figures
\ref{F5}, \ref{F6} and \ref{F7}. It can be observed that XGBoost
performs worse when only considering CN, A3 and L3 alone, although
good values of precision (CN:0.9916, A2:0.9807 and A3:0.9838) and
recall (CN:0.6704, A2:0.7012 and A3:0.7037) are achieved.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F5}Results for CN feature only}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA2_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA2_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F6}Results for A3 feature alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyA3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyA3_SUMMARY}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F7}Results for L3 feature alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Only_Metric/ROConlyL3_SUMMARY}\includegraphics[width=0.48\columnwidth]{Only_Metric/CMonlyL3_SUMMARY}
\end{figure}

Finally, an relevant evaluation on the models that include a handcrafted
feature is necessary: How important is the appended feature for the
model result? The answer can be observed by the feature importance
plots for CN, A3 and L3 in Figure \ref{F8-importance}. Each plot
resembles the importance gain from each feature in the underlying
decision trees that \texttt{XGBoost} uses. Usually, the more bifurcations,
the higher gain a feature has, the more relevance a feature has on
the model prediction itself. All models have the appended feature
as the most important by a good margin, although margin for A3 is
not wide as for A2 or L3.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
with each feature}
\par\end{centering}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
\par\end{raggedleft}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
\par\end{raggedleft}
\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
\end{figure}

@@

As the first step in the evaluation of CN, A3 and L3 methods for edge
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figures \ref{fig:HI1} and \ref{fig:HI2},
in which the x axis represents a rank position $k$ and the y axis
displays the precision for the top $k$ predictions of each method,
when assuming interactome \emph{HI-III} as the validation set.

\begin{figure}[h]
\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14}}

@@

\begin{figure}[h]
\caption{\label{fig:HI2}Methods Comparison for \emph{HI-TESTED}}

@@

However, it is of our interest to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
was adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
now the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned before. Results
for the Interactome \emph{HI-II-14} with Node2Vec and each metric
are presented in Figures \ref{F2-1}, \ref{F3-1} and \ref{F4-1}.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2-1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3-1}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4-1}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
\end{figure}

It can be seen that for \emph{HI-II-14}, L3 performs marginally better
than CN, especially when comparing precision ( vs ) and recall ( vs
), but all of them have good values for area under the ROC curve of
0.99.

@@

Based on previous work from other authors\cite{Kovacs2019}, the intrinsic
knowledge of existing protein-protein interaction networks let scientists
elucidate promising new interactions to explore. In this study, a
machine learning algorithm was fed with previously elucidated information
about the network, as well as results form a state-of-the-art edge
embedding technique. Even if the baseline of the handcrafted features
(CN, A3, L3) yields good results, boosting that information with the
vector embedding, which calculates vector representations of the network
structure improves the overall performance of each method on its own.
